Survival in chessland

### 2019-02-18 chess/papers/survival.tex
Rename Player to StatelessPlayer, in preparation for allowing stateful algorithms

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
  rules technically do not allow a draw until both players have made a
  move (5.2.3). There are always other routes to a draw, for example
  by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both players' desire. Since the rules of
  chess-to-the-death can't forbid us from colluding right now as you
  read this paper, I hereby declare that the following is the correct
  approach:
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] \ldots Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    % XXX potential citation to 'elo world' if players are stateful..
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] \ldots Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

  Since this is one of the shortest possible routes to a draw, I hereby
  dub this line the ``Fool's Draw,'' by analogy with the Fool's Mate.

@@

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience
were also consulted for their wagers; these are compared in
Section~\ref{sec:buddies}.

@@

To compute the chances for survival, I legally acquired 506,000,416
chess games from lichess.org. This is all of the standard variant,
rated games from Jan 2013 to November 2018, in any time format. Only
games that are completed and valid are included (about 200,000 games
did not meet this criteria). The total data size is 875 gigabytes, so
processing these took some care for efficiency and parallelism.

Other than that, I simply parsed and simulated each game. For each
of the 32 pieces in the starting position, I tracked its current
location, and whether it is alive; multiple dead pieces could
occupy the same square. At the end of the game, one of the kings is
killed if his side has lost.

@@

\begin{figure}[htb]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
per side) and bullet ($\sim 1$ minute) games are predominant. In
Section~\ref{sec:slices} I show some slices of the data for
comparison. % XXX could summarize whether this is a real effect or
% not, once I have the slices!

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players, although there are a few things that
% are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

\section{Guesses and slices}

@@

It is a good and humbling act to compare the actual results to the
hypotheses gathered before the experiment. Figure~\ref{fig:buddies}
compares the ranking across all games (slice {\bf Actual}) to the
author author (slice {\bf Tom}) and his drinking buddies (slices {\bf
  Ben}, {\bf Jim}, {\bf David}, {\bf William}). There are a number of
different reasonable ways to measure the accuracy of this type of
position; a very simple one is the sum of the absolute differences in
rank for each piece (e.g. if in one ranking \king is \#3, and in the
other \#5, then this contributes 2 to the total error). By this
metric, Ben has the best prediction (98 error), followed by David
(138) and Tom (148). Tom and David had the most similar predictions
(116) and Jim and William the most different (312). The expected error
between two completely random permutations is about 341, so all of
these guesses are significantly better than chance. Note in the actual
ranking, many pieces have very similar survival probabilities, and
many guesses are ambivalent about groups of pieces. Weighting each
rank difference equally is therefore an oversimplification. It would
have been better to ask each participant to give probabilities, as
David did; this would give us more sensitive error metrics and more
opportunities to spend the afternoon making visualizations.

@@

\subsection{Slices} \label{sec:slices}
The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} slice in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master'' title
  to notable players on the site; this title is excluded.}). These
games have high-quality play, although there are not many of them
(only XXX), so the variance is somewhat higher. % XXX give error bounds?

The other slices, {\bf bullet}, {\bf blitz}, {\bf rapid} and {\bf
  classical} are different time control formats. Games on Lichess are
played with a starting clock (per side) and an increment added to the
clock after each player's move. The total time is taken to be the
starting time plus 40 times the increment (with the idea that an
average game has 40 moves per side); this is the same formula that
Lichess uses. A bullet game is when the total time per side is
between 30~seconds and 3~minutes; blitz is between 3 and 8~minutes;
rapid is between 8 and 25~minutes; classical is any more than this
(including untimed games).

>>>>

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves), or without the arbiter's consent, and FIDE
  rules technically do not allow a draw until both players have made a
  move (5.2.3). There are always other routes to a draw, for example
  by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both players' desire. Since the rules of
  chess-to-the-death can't forbid us from colluding right now as you
  read this paper, I hereby declare that the following is the correct
  approach:
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] \ldots Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players\cite{eloworld} may simply copy what white does.
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] \ldots Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

  Since this is one of the shortest possible routes to a draw, I
  hereby dub this line the ``Fool's Draw,'' by analogy with the Fool's
  Mate. In this line all pieces survive, which is anyway humane and
  also advantageous in the case that you are someone is simultaneously
  being one of the chesspieces to the death!

@@

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience
were also consulted for their wagers; these are compared to each other
and to the actual results in Section~\ref{sec:guesses}.

@@

To compute the chances for survival, I legally acquired 506,000,416
chess games from lichess.org. This is all of the standard variant,
rated games from Jan 2013 to November 2018, in any time format. Only
games that are completed and valid are included (about 200,000 games
did not meet this criteria). The total data size is 875 gigabytes, so
processing these took some care for efficiency and parallelism.
Fortunately, I have a computer with just an obscene number of cores
and truly excessive RAM, so you gotta use that for
something.\footnote{ To torture your own desktop computer, source code
  is available at
  \url{sourceforge.net/p/tom7misc/svn/HEAD/tree/trunk/chess/}.}

Other than that, I simply implemented the rules of chess, wrote a PGN
parser, then parsed and simulated each game. For each
of the 32 pieces in the starting position, I tracked its current
location, and whether it is alive; multiple dead pieces could
occupy the same square. At the end of the game, one of the kings is
killed if his side has lost.

@@

\begin{figure}[t]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim\! 5$
minutes per side) and bullet ($\sim\! 1$ minute) games are
predominant. Nonetheless, the results are fairly robust across
different time formats and skill levels. In Section~\ref{sec:slices} I
show some slices of the data for comparison.

@@

% XXX using this a way to play the game

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players (in new dataset too), although there are
% a few things that are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

@@

\section{Guesses and slices} \label{sec:guesses}

@@

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is a
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

@@

\subsection{Slices} \label{sec:slices}

\begin{figure}
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}

The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} slice in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
  title to notable players on the site; this title is excluded from
  the sample.}). These games have high-quality play, but far fewer
samples (``only'' 3.4 million). This set exhibits significant
variance; for example, the survival rate of \pawn c ranges from
$38$--$46\%$. This is both because of the small sample size and the
bucketization by player name; there are few enough titled players that
an individual's preference in openings and style of play moves the
values of their entire bucket. I caution against reading too much into
this column. It seems we can at least conclude that these games tend
to be much bloodier (Kings aside, survival rates are lower across the
board); top players are less likely to fall for traps early in the
game, perhaps more willing to sacrifice material, and more likely to
play into endgames where almost every piece is exchanged. If being a
chesspiece to the death, you do {\em not} want to have a Grandmaster
playing the game!

% XXX give % error % bounds?

On the other hand, the other slices all have enough samples that the
variance is minimal. These slices, {\bf bullet} (151,261,707 games),
{\bf blitz} (236,050,938 games), {\bf rapid} (98,606,558) and {\bf
  classical} (13,886,352) are different time control formats. Games on
Lichess are played with a starting clock (per side) and an increment
added to the clock after each player's move. The game is classified
according to the estimated total time: The starting time plus $40 \times$
the increment (with the idea that an average game has 40 moves per
side); this is the same formula that Lichess uses. A bullet game is
when the total time per side is between 30~seconds and 3~minutes;
blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
classical is any more than this (including untimed games). Each of
these slices has enough samples that the variance is very low.
Here we see that the ranking is rather stable across the range of
time formats, which was not what I expected. This should increase
our confidence that the results are really inherent to chess, not
just the particulars of this data set.
